---
title: "Using Decorators in mcpr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using Decorators in mcpr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(mcpr)
```

## Introduction

The mcpr package provides a decorator system similar to plumber for defining MCP (Model Context Protocol) servers. Decorators allow you to annotate R functions with special comments that describe how they should be exposed through MCP. This approach is particularly useful when you have existing R code that you want to make available to AI assistants like Claude.

## Decorator Syntax

Decorators in mcpr use the `#*` prefix (similar to plumber) followed by decorator tags. The basic structure is:

```r
#* @decorator_type
#* @tag value
#* @another_tag another value
function_name <- function(parameters) {
  # Function implementation
}
```

## Available Decorators

### @mcp_tool

The `@mcp_tool` decorator exposes a function as an MCP tool that can be called by AI assistants.

```r
#* @mcp_tool
#* @description Calculate the mean of a numeric vector
#* @param x numeric vector to analyze
#* @param na.rm logical whether to remove NA values (default: TRUE)
calculate_mean <- function(x, na.rm = TRUE) {
  mean(x, na.rm = na.rm)
}
```

**Required tags:**
- `@description` - A clear description of what the tool does

**Optional tags:**
- `@param` - Parameter documentation (format: `@param name description`)
- `@returns` - Description of the return value

### @mcp_resource

The `@mcp_resource` decorator exposes a function as an MCP resource that provides data or information.

```r
#* @mcp_resource
#* @description Get system information
#* @mime_type application/json
get_system_info <- function() {
  list(
    r_version = R.version.string,
    platform = R.version$platform,
    locale = Sys.getlocale(),
    timezone = Sys.timezone()
  )
}
```

**Required tags:**
- `@description` - Description of the resource

**Optional tags:**
- `@mime_type` - MIME type of the resource (default: "application/json")

### @mcp_prompt

The `@mcp_prompt` decorator defines a prompt template that AI assistants can use.

```r
#* @mcp_prompt
#* @description Generate a data analysis report
#* @template Analyze the {{dataset}} using {{method}} and focus on {{aspect}}. 
#*   Provide visualizations and statistical insights.
#* @param_dataset The dataset to analyze
#* @param_method The analysis method (e.g., "regression", "clustering")
#* @param_aspect The aspect to focus on (e.g., "trends", "outliers")
data_analysis_prompt <- NULL
```

**Required tags:**
- `@description` - Description of the prompt
- `@template` - The prompt template with {{parameter}} placeholders

**Optional tags:**
- `@param_*` - Parameter descriptions (format: `@param_name description`)

## Parameter Type Annotations

You can specify parameter types in the `@param` documentation:

```r
#* @mcp_tool
#* @description Perform statistical analysis
#* @param data numeric vector or data.frame to analyze
#* @param method character analysis method ("mean", "median", "sd")
#* @param conf.level numeric confidence level (0-1)
#* @param plot logical whether to create a plot
analyze_data <- function(data, method = "mean", conf.level = 0.95, plot = FALSE) {
  # Implementation
}
```

### Type Mappings

mcpr automatically maps R types to JSON Schema types:

| R Type | JSON Schema Type | Example |
|--------|------------------|---------|
| numeric | number | `@param x numeric values` |
| integer | integer | `@param n integer count` |
| character | string | `@param name character string` |
| logical | boolean | `@param flag logical TRUE/FALSE` |
| list | object | `@param options list of settings` |
| data.frame | array | `@param df data.frame dataset` |

## Using Decorated Functions

### Loading from a File

Use the `mcp_source()` method to load decorated functions:

```r
# Create a server
server <- mcp(name = "My Analysis Server", version = "1.0.0")

# Load decorated functions from a file
server$mcp_source("analysis-tools.R")

# Run the server
server$mcp_run(transport = "http", port = 8080)
```

### Loading from a Directory

Use `mcp_source_directory()` to load all R files with decorators:

```r
# Load all decorated functions from a directory
server$mcp_source_directory("R/", recursive = TRUE)
```

### Scanning for Decorated Files

Use `scan_mcp_directory()` to find files containing MCP decorators:

```r
# Find all files with MCP decorators
decorated_files <- scan_mcp_directory("R/")
print(decorated_files)
```

## Complete Example

Here's a complete example file with various decorator types:

```r
# statistics-tools.R

#* @mcp_tool
#* @description Calculate descriptive statistics for a numeric vector
#* @param x numeric vector to analyze
#* @param na.rm logical whether to remove NA values (default: TRUE)
#* @param digits integer number of decimal places (default: 2)
describe <- function(x, na.rm = TRUE, digits = 2) {
  if (!is.numeric(x)) {
    stop("x must be numeric")
  }
  
  stats <- list(
    n = length(x),
    mean = mean(x, na.rm = na.rm),
    median = median(x, na.rm = na.rm),
    sd = sd(x, na.rm = na.rm),
    min = min(x, na.rm = na.rm),
    max = max(x, na.rm = na.rm),
    q1 = quantile(x, 0.25, na.rm = na.rm),
    q3 = quantile(x, 0.75, na.rm = na.rm)
  )
  
  # Round to specified digits
  lapply(stats, round, digits = digits)
}

#* @mcp_tool
#* @description Create a frequency table for categorical data
#* @param x character or factor vector
#* @param sort logical whether to sort by frequency (default: TRUE)
#* @param prop logical whether to include proportions (default: TRUE)
frequency_table <- function(x, sort = TRUE, prop = TRUE) {
  tbl <- table(x)
  
  if (sort) {
    tbl <- sort(tbl, decreasing = TRUE)
  }
  
  result <- data.frame(
    value = names(tbl),
    count = as.numeric(tbl),
    stringsAsFactors = FALSE
  )
  
  if (prop) {
    result$proportion <- result$count / sum(result$count)
    result$percentage <- round(result$proportion * 100, 1)
  }
  
  result
}

#* @mcp_resource
#* @description Get available example datasets
#* @mime_type application/json
list_example_datasets <- function() {
  # Get datasets from base R
  data_list <- data(package = "datasets")$results
  
  datasets <- data.frame(
    name = data_list[, "Item"],
    title = data_list[, "Title"],
    stringsAsFactors = FALSE
  )
  
  # Add some metadata
  datasets$rows <- NA
  datasets$cols <- NA
  datasets$class <- NA
  
  for (i in 1:nrow(datasets)) {
    dataset_name <- datasets$name[i]
    if (exists(dataset_name)) {
      obj <- get(dataset_name)
      if (is.data.frame(obj) || is.matrix(obj)) {
        datasets$rows[i] <- nrow(obj)
        datasets$cols[i] <- ncol(obj)
      }
      datasets$class[i] <- class(obj)[1]
    }
  }
  
  datasets
}

#* @mcp_resource
#* @description Load an example dataset
#* @mime_type application/json
load_dataset <- function(name) {
  if (!exists(name)) {
    stop(paste("Dataset", name, "not found"))
  }
  
  dataset <- get(name)
  
  # Convert to list format for JSON serialization
  if (is.data.frame(dataset)) {
    return(as.list(dataset))
  } else if (is.matrix(dataset)) {
    return(list(data = dataset, dim = dim(dataset)))
  } else {
    return(dataset)
  }
}

#* @mcp_prompt
#* @description Request a statistical analysis
#* @template Please analyze the {{dataset}} dataset focusing on {{variables}}.
#*   Use {{methods}} and provide {{outputs}}.
#*   
#*   Additional requirements: {{requirements}}
#* @param_dataset The name of the dataset to analyze
#* @param_variables The variables to focus on (comma-separated)
#* @param_methods The statistical methods to apply
#* @param_outputs The desired outputs (e.g., "summary statistics, plots, tests")
#* @param_requirements Any additional analysis requirements
statistical_analysis_request <- NULL

#* @mcp_prompt
#* @description Request a data visualization
#* @template Create a {{plot_type}} visualization for the {{dataset}} dataset.
#*   Show the relationship between {{x_var}} and {{y_var}}.
#*   {{additional_instructions}}
#* @param_plot_type The type of plot (e.g., "scatter", "bar", "histogram")
#* @param_dataset The dataset to visualize
#* @param_x_var The x-axis variable
#* @param_y_var The y-axis variable (if applicable)
#* @param_additional_instructions Any additional plotting instructions
visualization_request <- NULL
```

## Best Practices

1. **Clear Descriptions**: Always provide clear, concise descriptions for your decorators
2. **Parameter Documentation**: Document all parameters with their expected types
3. **Error Handling**: Include appropriate error handling in your functions
4. **Return Values**: Ensure your functions return JSON-serializable objects
5. **Naming Conventions**: Use descriptive, consistent names for your functions

## Advanced Features

### Multi-line Descriptions

Use indentation for multi-line descriptions:

```r
#* @mcp_tool
#* @description Perform comprehensive statistical analysis including:
#*   - Descriptive statistics
#*   - Normality tests
#*   - Correlation analysis
#*   - Basic visualizations
comprehensive_analysis <- function(data) {
  # Implementation
}
```

### Excluding Functions

While there's no `@mcp_ignore` decorator yet, you can simply not decorate functions you don't want to expose:

```r
# This function will be exposed
#* @mcp_tool
#* @description Public analysis function
analyze_public <- function(data) {
  process_internal(data)
}

# This function will NOT be exposed (no decorator)
process_internal <- function(data) {
  # Internal processing
}
```

## Integration with Server Generation

Decorated functions work seamlessly with mcpr's server generation features:

```r
# Generate a standalone server from decorated files
generate_mcp_server(
  name = "stats-server",
  source = "statistics-tools.R",
  output_dir = "./servers"
)
```

## Troubleshooting

### Common Issues

1. **Functions not appearing**: Ensure you're using `#*` (not just `#`) for decorators
2. **Parameter types not recognized**: Check that parameter names in `@param` match function arguments
3. **JSON serialization errors**: Ensure return values are JSON-serializable
4. **File not loading**: Verify the file path and that the file contains valid R code

### Debugging

To see parsed decorator information:

```r
# Parse decorators without registering
parsed <- parse_mcp_decorators("my-file.R")
str(parsed)
```

## Conclusion

The decorator system in mcpr provides a clean, intuitive way to expose R functions through MCP. By using familiar roxygen2-style comments, you can quickly make your R code available to AI assistants while maintaining readable, well-documented code.

For more examples, see the `inst/examples/decorated-functions.R` file in the mcpr package.